Human capital drives natural selection in contemporary humans

### 2021-04-14 why-natural-selection.Rmd
Address comments in PNAS review - new title - Use standard_ggplot more - Add confidence intervals - Main plot to start - Discuss endogeneity - Add computation of CARA utility

---
title: "Natural selection in contemporary humans is driven by correlations with human capital"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{xr-hyper}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

theme_set(theme_minimal())
my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "#FAC32D")

standard_ggplot <- function (
        dfr, 
        fill_col, 
        n_regs, 
        ...,
        score_col = quo(term), 
        order_idx = 1, 
        n_cats    = NULL
      ) {
  if (! missing(score_col)) score_col <- enquo(score_col)
  mfc <- missing(fill_col)
  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
  if (missing(n_cats)) {
    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
  }

@@

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        rename(p = p.value) %>% 
        ggplot(aes(estimate, {{score_col}}, fill = {{fill_col}}, ...,
          colour = p < 0.05/{{n_regs}})) +
          pval_geom_point() +
          my_vline + 
          pval_scale() +
          my_fill_scale(n = n_cats) +
          my_labs() +
          my_theme()
}

@@

pval_geom_point <- function (size = 1.5, stroke = 0.5, ...) {
  geom_point(shape = 21, size = size, stroke = stroke, ...)
}

my_fill_scale <- function (n, aesthetics = "fill", ...) {
  if (n == 2) {
    scale_colour_manual(values = c("steelblue4", "#D63C3C"), 
          aesthetics = aesthetics, ...)
    # scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

@@

```{r NOTES-TODO, eval=FALSE}
"
# Open questions/TODO

@@

To test directly for natural selection, we regress respondents' number of
children ($y_i$) on their polygenic scores (PGS):

```{=tex}
\begin{equation}
y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

@@

[^normalization]: The selection effect $\beta$ equals $Cov(Y, PGS)/Var(PGS)$
where $Y$ is the number of children. Since PGS are normalized to variance 1,
this reduces to to $Cov(Y, PGS) \equiv E(Y PGS) - E(Y) E(PGS)$, which in turn
reduces to $E(Y PGS)$ as $E(PGS) = 0$. This is the polygenic score weighted by
the number of children, which is the average polygenic score in the next
generation.

@@

Since UK Biobank participants are not representative of the wider population,
naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using population
data. We try three alternative weighting schemes: weighting by geography, age
and presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes.
Mean effect sizes across all polygenic scores are
increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables.[^purifying]

[^purifying]: We also check for purifying selection by estimating \@ref(eq:regression) with a quadratic term. See Appendix Figure \@ref(fig:plot-purifying).

@@

To understand the reasons for this ascertainment bias, and to learn about the
mechanisms underlying natural selection, we split the sample, starting with
basic demographic variables including education, income and sex. These are all
potential sources of ascertainment bias, as mentioned above.

@@

res_edu %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, colour = p < 0.05/{{n_regs}}, 
        fill = `Age left FTE`)) +
        my_vline +
        pval_geom_point() + 
        pval_scale() +
        my_fill_scale(n = 3, direction = -1) +
        my_labs() +
        my_theme()

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs(1)),
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        ),
        Income = add_n(Income, "income_cat")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = Income, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point(alpha = 0.8) + 
        my_fill_scale(n = 5, direction = -1) +
        pval_scale() +
        my_labs() +
        my_theme()
```

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. Note that the overall
effect is not a simple average of the effect among the different subgroups,
because polygenic scores may also shift respondents between the subgroups. For
example, a high polygenic score for educational attainment may predict having
fewer children among early school leavers, but it may also increase the age at
which a respondent leaves school.

@@

These results could be explained by age, if older respondents have lower income
and are less educated, and also show more natural selection on polygenic
scores. However, if we rerun the regressions, interacting the polygenic
score with income category and also with a quadratic in age, the
interaction with income remains significant at 0.05/`r n_regs_ic` for
`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
the PGS with age of leaving full time education and a quadratic in age,
the interaction with age leaving full time education remains significant at
0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.

Selection effects are also different between men and women (Appendix Figure
\@ref(fig:plot-sexes)). Differences are particularly large for educational
attainment, height and MDD. Several polygenic scores for mental illness and
personality traits are more selected for (or less against) among women,
including major depressive disorder (MDD), schizophrenia and neuroticism, while
extraversion is more selected for among men. Scores for waist circumference and
waist-hip ratio are less selected for among women, and scores for educational
attainment are more selected against, though the difference is only significant
for the ("EA3") educational attainment score of @lee2018gene. One
possible reason for these sex differences is that polygenic scores may affect
fertility via success in marriage markets, and men and women may value different
characteristics in these.

@@

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(-5, 0, 0, -85),
                  legend.margin        = margin(0, 0, 0, 0),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.spacing       = unit(1.5, "lines")
                )
sexes_guides <- guides(
                  # fill  = guide_legend(ncol = 1),
                  # color = guide_legend(ncol = 1)
                )

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        sexes_guides

@@

res_with_partner_sex %>% 
      mutate(
        score_name = pretty_names(score_name),
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = Household, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        sexes_guides

@@

res_age_flb_cross %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(1)),
        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
        term       = add_n(term, "age_flb_cat")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, score_name, fill = term, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 3, direction = -1, 
          guide = guide_legend(title = "Age at first live birth")) +
        my_labs() +
        my_theme()

@@

res_combined %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Uncontrolled", "Controlled for AFLB")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p < 0.05/{{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()
```

@@

Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
number of children and number of grandchildren (for the parents' generation). We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-17}$).

## Economic fertility theory and natural selection

@@

```{r, eval = FALSE}
"Other models of the literature

@@

Another alternative is that traits under selection are linked to externalizing
behaviour and risk-seeking. This might be captured by our risk-aversion 
parameter $\sigma$, which affects fertility by changing the marginal utility of income, but a more direct channel is risky sexual behaviour
[@Mills2020.05.06.081273]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
(younger) age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. This non-result could simply
reflect the imprecision of the risk aversion measure, which is a single yes/no
question. However, the measure *does* predict the overall number of children,
highly significantly ($p \le 2^{-16}$, `r n_risk_sig`/33 regressions). Given that,
and our large N, we believe that the non-result is real: while risk aversion does
predict fertility in the sample, it is not an important channel for natural 
selection.

@@

Existing evidence on human natural selection has led some to "biocosmic
pessimism" [@sarraf2019modernity]. Others are more sanguine, and argue that
natural selection's effects are outweighed by environmental improvements, such
as those underlying the Flynn effect [@flynn1987massive]. The evidence
here may add some nuance to this debate. Patterns of natural selection have been
relatively consistent across the past two generations, but they are not the
outcome of a universal, society-wide phenomenon. Instead they result from a sum
of opposing forces, operating in different parts of society and pulling in
different directions.

@@

```{r plot-sexes, fig.cap = "Selection effects by sex. Highlighted lines are significant differences. Highlighted points are significantly different from 0."}

@@

res_sex_wide <- res_sex %>% 
      select(term, estimate, sex, diff.p.value) %>% 
      pivot_wider(names_from = "sex", values_from = "estimate") %>% 
      mutate(
        min_est = pmin(Female, Male),
        max_est = pmax(Female, Male),
        term = pretty_names(term),
        term = fct_reorder(term, Female)
      )

res_sex %>% 
      mutate(
        term = pretty_names(term),
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p < 0.05/{{n_regs}})) +
        my_vline +
        geom_linerange(data = res_sex_wide, 
          aes(
            y      = term,
            x      = NULL,
            fill   = NULL,
            xmin   = min_est, 
            xmax   = max_est, 
            colour = diff.p.value < 0.05/{{n_regs}}
          ), 
        group = "") +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()

@@

res_age_birth_parents %>% 
    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
    mutate(
      term    = pretty_names(term),
      term    = fct_reorder(term, estimate, order_abs(2)),
      Control = ifelse(control == "fath_age_birth", 
                  "Father's age at birth", 
                  "Mother's age at birth"
                )
    ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(y = term, x = estimate, fill = Control, 
      colour = p < 0.05/{{n_regs}})) + 
      pval_geom_point(alpha = 0.8) +  
      my_vline +
      my_fill_scale(2) +
      pval_scale() +
      my_labs() +
      my_theme()

@@

res_age_flb_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, term, colour = p < 0.05 / {{n_regs}})) + 
      my_vline +
      pval_geom_point(fill = "steelblue4") +
      pval_scale() + 
      my_labs() +
      my_theme()

@@

res_age_birth_parents_dv %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1)),
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's AFLB", 
          "moth_age_birth" = "Mother's AFLB"
        )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(estimate, term, fill = `Dependent variable`, 
        colour = p < 0.05 / {{n_regs}})) + 
        my_vline +
        pval_geom_point() +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()

>>>>

---
title: "Human capital drives natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

theme_set(theme_minimal())

my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "#FAC32D")

standard_ggplot <- function (
        dfr, 
        fill_col, 
        n_regs, 
        ...,
        score_col = quo(term), 
        order_idx = 1, 
        fill_direction = 1,
        n_cats    = NULL,
        conf_int  = NULL
      ) {
  if (! missing(score_col)) score_col <- enquo(score_col)

  mfc <- missing(fill_col)
  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
  if (missing(n_cats)) {
    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
  }

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

@@

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        rename(p = p.value) %>% 
        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
          my_vline + 
          confint_geom_segment +
          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
          pval_scale() +
          my_fill_scale(n = n_cats, direction = fill_direction) +
          my_labs() +
          my_theme() +
          scale_linetype_manual(name = "", values = c("95% c.i." = 1))

}

@@

pval_geom_point <- function (size = 1.5, stroke = 0.5, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i."), 
                    color = "grey60", alpha = 0.25, size = 1.1, 
                    lineend = "round")
}

my_fill_scale <- function (n, aesthetics = "fill", ...) {

  fill_colors <- c("steelblue4", "#D63C3C")

  if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

@@

```{r NOTES-TODO, eval=FALSE}
"

# TODO response to PNAS

* Add CIs - done.
* Clarify collider bias/endogeneity - done.
* Formalize theory - done.
  - Comment on results where effects are actually reversed. Could this
  happen if \sigma were higher in period 2? Or if it increased with income?
  Could maybe do a computational example.
  - One possibility is CARA utility. Solving it computationally
* Clarify causality?
  - Helps to deal with the "collider bias" argument of reviewer
  - Might also be a better test of mediation via human capital
  - One other possibility. If we use only people with first profession
    teacher, that (as it happens) shuts down the EA3-income correlation.
    It also shuts down the EA3-fertility relationship, suggesting that
    income is indeed the mediator.
* Control for PGS for fertility
* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
  - Point is that we can explain how natural selection is occurring.
* "Discuss empirical predictions for AFLB grounded in dynamic theory"
  - hmm, tricky, maybe we can include it in discussion of economic results at
  end of formal section.
* Why don't women see stronger effects? But note, they do: not significant
  after correction for multiple testing, but a 25% difference.

# Open questions/TODO

@@

To test directly for natural selection, we regress respondents' number of
children ($N_i$) on their polygenic scores (PGS):

```{=tex}
\begin{equation}
N_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
\label{eq:regression}
\end{equation}
```

@@

[^normalization]: The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
where $N$ is the number of children. Since PGS are normalized to variance 1 and
mean 0, this reduces to 
$Cov(N, PGS) \equiv E(N\ PGS) - E(N) E(PGS) = E(N\ PGS)$. This is the polygenic 
score weighted by the number of children, which is the average polygenic score 
in the next generation.

@@

Since UK Biobank participants are not representative of the wider population,
naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using population
data. We try three alternative weighting schemes: weighting by geography, age
and presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes.
Mean effect sizes across all polygenic scores are
increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables.

We also check for purifying and diversifying selection by 
estimating \@ref(eq:regression) with a quadratic term. See Appendix Figure \@ref(fig:plot-purifying). In particular, we find diversifying selection for
educational attainment polygenic scores: at higher values of these scores, 
the effect on fertility is closer to zero.

@@

To understand the reasons for this ascertainment bias, and to learn about the
mechanisms underlying natural selection, we split the sample, starting with
basic demographic variables including education, income and sex. These are all
potential sources of ascertainment bias, as mentioned above. We then re-estimate
\@ref(eq:regression) within each subgroup.

We emphasize two points. First, this exercise does
*not* apportion the total correlation with fertility between subgroups. 
That is because the polygenic score may also predict the probability of being in 
each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
fertility among university graduates, and among non-graduates; it also changes
a person's chance of attending university. Second, we do *not* claim that our
observed correlations reflect causal effects of polygenic scores on fertility.
Indeed, polygenic scores correlate with many sources of environmental variation 
that could affect fertility. However, natural selection is a correlational
phenomenon, not a causal one. That is, if a polygenic score correlates with 
low fertility, it is being selected against, whatever the underlying causal
relationship. Differences in equation \@ref(eq:regression) across subgroups can 
be informative about how natural selection works, because they constrain the set 
of possible explanations to those that generate these differences.

@@

res_edu %>% 
      mutate(
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
      ) %>% 
      standard_ggplot(n_regs = n_regs, fill_col = `Age left FTE`, 
                        fill_direction = -1)

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        ),
        Income = add_n(Income, "income_cat")
      ) %>% 
      standard_ggplot(n_regs = n_regs, fill_col = `Income`, 
                        fill_direction = -1)
```

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category.

@@

These results could be explained by age, if older respondents have lower income
and are less educated, and also show more natural selection on polygenic
scores. However, when we rerun the regressions, interacting the polygenic
score with income category and also with a quadratic in age, the
interaction with income remains significant at 0.05/`r n_regs_ic` for
`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
the PGS with age of leaving full time education and a quadratic in age,
the interaction with age leaving full time education remains significant at
0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.

Selection effects are also different between men and women (Appendix Figure
\@ref(fig:plot-sexes)). Differences are particularly large for educational
attainment, height and MDD. Several polygenic scores for mental illness and
personality traits are more selected for (or less against) among women,
including major depressive disorder (MDD), schizophrenia and neuroticism, while
extraversion is more selected for among men. Scores for waist circumference and
waist-hip ratio are less selected for among women. One
possible reason for these sex differences is that polygenic scores may affect
fertility via success in marriage markets, and men and women may value different
characteristics in these.

@@

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(-5, 0, 0, -85),
                  legend.margin        = margin(0, 0, 0, 0),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.spacing       = unit(1, "lines")
                )
sexes_guides <- guides(
                  # fill  = guide_legend(ncol = 1),
                  # color = guide_legend(ncol = 1)
                )

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
        sexes_guides

@@

res_with_partner_sex %>% 
      mutate(
        score_name = pretty_names(score_name),
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = Household, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
        sexes_guides

@@

res_age_flb_cross %>% 
      mutate(
        `Age at first live birth` = gsub("age_flb_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, "age_flb_cat")
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs,
                      fill_direction = -1)

@@

res_combined %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Uncontrolled", "Controlled for AFLB")
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = Regression, n_regs = n_regs)

@@

Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
number of children and number of grandchildren (for the parents' generation). We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

## Human capital and natural selection

@@

Our results are actually stronger than this, in that correlations with
fertility are reversed at higher AFLB. This could occur if relative risk 
aversion increased at higher income levels.

```{r literature, eval = FALSE}
"Other models of the literature

@@

Another alternative is that traits under selection are linked to externalizing
behaviour and risk-seeking. This might be captured by our risk-aversion 
parameter $\sigma$, which affects fertility by changing the marginal utility of income, but a more direct channel is risky sexual behaviour
[@Mills2020.05.06.081273]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
(younger) age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. This non-result could simply
reflect the imprecision of the risk aversion measure, which is a single yes/no
question. However, the risk aversion measure *does* predict the overall number
of children, highly significantly ($p \le 2 \times 10^{-16}$, `r n_risk_sig`/33
regressions). Given that, and the statistical power we get from our large N, we 
believe that the non-result is real: while risk aversion does predict fertility 
in the sample, it is not an important channel for natural selection.

@@

Existing evidence on human natural selection has led some to "biocosmic
pessimism" [@sarraf2019modernity]. Others are more sanguine, and argue that
natural selection's effects are outweighed by environmental improvements, such
as those underlying the Flynn effect [@flynn1987massive]. The evidence
here may add some nuance to this debate. Patterns of natural selection have been
relatively consistent across the past two generations, but they are not the
outcome of a universal, society-wide phenomenon. Instead they result from  
opposing forces, operating in different parts of society and pulling in
different directions.

@@

```{r plot-sexes, fig.cap = "Selection effects by sex. Highlighted lines are significant differences at p < 0.05/33. Highlighted points are significantly different from 0 at p < 0.05/66."}

@@

res_sex_wide <- res_sex %>% 
      select(term, estimate, sex, diff.p.value) %>% 
      pivot_wider(names_from = "sex", values_from = "estimate") %>% 
      mutate(
        min_est = pmin(Female, Male),
        max_est = pmax(Female, Male),
        term = pretty_names(term),
        term = fct_reorder(term, Female)
      ) 
n_comps <-as.double(nrow(res_sex_wide))

res_sex %>% 
      mutate(
        term = pretty_names(term),
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = sex, 
            colour = p < 0.05/{{n_regs}})) +
        my_vline +
        geom_linerange(data = res_sex_wide, 
          aes(
            y      = term,
            x      = NULL,
            fill   = NULL,
            xmin   = min_est, 
            xmax   = max_est, 
            colour = diff.p.value < 0.05/{{n_comps}}
          ), 
        group = "") +
        pval_geom_point() +
        pval_scale(name = "p < 0.05 (corrected)") +
        my_fill_scale(n = 2) +
        my_labs() +
        my_theme()

@@

res_age_birth_parents %>% 
    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
    mutate(
      Control = ifelse(control == "fath_age_birth", 
                  "Father's age at birth", 
                  "Mother's age at birth"
                )
    ) %>% 
    standard_ggplot(fill_col = Control, n_regs = n_regs, order_idx = 2)

@@

res_age_flb_dv %>% standard_ggplot(n_regs = n_regs)

@@

res_age_birth_parents_dv %>% 
      mutate(
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's AFLB", 
          "moth_age_birth" = "Mother's AFLB"
        )
      ) %>% 
      standard_ggplot(fill_col = `Dependent variable`, n_regs = n_regs)
